---
title: "AE reporting"
author: "Eamonn O'Brien<br><small><br>Biostatistics<br></small>"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    use_bookdown: true
    toc_depth: 3
    fig_caption: true
    code_folding: hide
link-citations: yes
description: "xxxxxx"
---
<style>
.table {
	width: 50%;
}
p.caption {
  font-size: 0.8em;
  color: DarkBlue;
}
body { 
	max-width: 1900px; 
	margin: 0 auto !important; 
	margin-right: 0 !important;
	float: none !important; 
}
</style>

```{css, echo=FALSE}

#https://designers.hubspot.com/docs/snippets/design/centering-your-website-using-max-width-and-auto-margins
#https://github.com/juba/rmdformats/issues/49
#https://stackoverflow.com/questions/60176526/kableextra-change-font-size-of-table-footnote
#https://towardsdatascience.com/create-a-word-cloud-with-r-bde3e7422e8a
 
```


```{r setup,echo=FALSE,include=FALSE}

  set.seed(234)
  library(rms)          # automatically engages rms(Hmisc)
  library(tidyverse)
  library(kableExtra) 
  library(knitr) 
  # wordcloud
  library(wordcloud)
  library(wordcloud2)
  library(RColorBrewer)
  library(tm)
  
  knitrSet(lang='markdown', w=7, h=7, dev='svg', fig.path='folder/')
  options(prType='html')
  options(width=200)
  
  # format function 
     formatz <- function(x){
        if (!is.na(x)  ) {
          sprintf(x, fmt = '%#.1f')  
         }
   }
 
```

# Introduction

Simple AE table example using kable in Rmarkdown

```{r, }

  m    <- 10      # m SOC categories we create, each will have m preferred terms
  n    <- 1500    # total sample size
  rate <- 0.33    # AE poisson rate
  J    <- 20      # maximum number of preferred terms that can occur

  
```

#  Make a fabricated medical dictionary

```{r, }

  require(stringi)
  require(word.alignment)

  myFun <- function(free="SOC.",n=1) { 
    
    paste0(free,paste(stri_rand_strings(sample(4:8,1),   # no of words
                                        sample(4:12,1)), # no of letters in words
                      #pattern = "[A-Za-z]",
                      collapse= " "))
  }
  
  N   <-  m

  x   <-  replicate(N, myFun()) # create N SOC groups
  p   <-  round(runif(N,2,J))   # size of SOC groups
  SOC <-  sort(rep(x, times=p)) # fill out the SOCs
  PT  <-  replicate(length(SOC), myFun(free="PT."))  # create unique Preferred terms 
  SOC <-  str_to_sentence(SOC)   # changes case so only first letter is uppercase
  PT <-   str_to_sentence(PT)    # changes case so only first letter is uppercase
  PT <- gsub("Pt","PT",PT)       # change case back
  SOC <- gsub("Soc","SOC",SOC)   # change case back
  db  <-  data.frame(cbind(SOC,PT))  # bind together, we now have the dictionary!   
  
```

#  Create some AE data

```{r, }

 id <- 1:n             # patients
 ae <- rpois(n, rate)  # AEs per person
 
 zeroAE <- id[ae %in% 0]  # these ids have no AEs
 elements_2_remove <- zeroAE
 
 people.with.AE =   id[!(id %in% elements_2_remove)]
 people.with.no.AE = id[(id %in% elements_2_remove)]
 
 AE <- ae[!ae %in% 0]
 
 ex <- rep(people.with.AE, AE) 
 
 d1 <- data.frame(cbind(id=ex))
 d1 <- add_count(d1,id)
 
 d2 <- d1 %>%
   group_by(id, n) %>%
   mutate(count = seq(n()))

 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~#people with no AEs
 people.with.no.AE = id[(id %in% elements_2_remove)]
 
 AE <- ae[ae %in% 0]
 
 d1 <- data.frame(cbind(id=people.with.no.AE, n=0,count=0))
 
 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 all <- rbind(d1,d2)
 
 all <- plyr::arrange(all, id, count)
 
 names(all) <- c("id","tot.AE","count.AE")

 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  AES <- db[sample(nrow(db), nrow(d2), replace=TRUE), ]
 
  d2 <- cbind(d2, AES)
 
 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  d1$SOC="";d1$PT=""
  
  all <- rbind(d1,d2)
  
  all <- plyr::arrange(all, id, count)
  
  names(all) <- c("id","tot.AE","count.AE","SOC","PT")
 
  #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  # add trt
  trt <- sample(c("Treatment A","Treatment B"), n, replace=TRUE)
  
  all$treatment <- rep(trt, times=as.vector(table(all$id)))
  all$treatment <- factor(all$treatment)

```

#  Calculate N

```{r, }  
  
  N <- unique(all[, c("id", "treatment")])
  N0 <- addmargins(table(N$treatment))
 
```

# Calculate the number of patients with at least one AE

```{r, }    
  
  N <- unique(all[, c("id", "tot.AE","treatment")])
  atleast <- N[!N$tot.AE %in% 0,]
  N1 <- addmargins(table(atleast$treatment))

```

# Manage data for by System order class : Preferred terms, the right way

```{r, eval=TRUE}  
 
  
  call <- all[!all$count.AE %in% 0,]  # dispense with all obs with no AEs
  
  s <- call[, c("id","SOC","PT","treatment")]  # drop count vars
  
 # s <- unique(s)  # if within a patient has multiple SOC and linked PTs, 
                  # keep only 1 occurrence as this extra info is not used
                  # this will not make a difference as I deal with this in the following code
  
  # s<-s[!(s$SOC=="" ),]  # no need for this as 
  
  fooz<-list()                        # this will be used to collect data
  
  # analyse data in SOC chunks 

  k <- sort(unique(s$SOC))       # get a vector of unique SOC
  L <- length(k)                 # how many unique SOC 
  
  for (i in 1:L) {               # one by one analyze SOC groups
   
    ss <- s[s$SOC %in% k[i],]    # isolate a SOC chunk
 
       # special case if only 1 SOC row
        if (nrow(ss) %in% 1) {
          
            ss1 <- addmargins(table(ss$PT, ss$treatment))
            Tot <- ss1[rownames(ss1) %in% "Sum",]
            body <- ss1[!rownames(ss1) %in% "Sum",]
            z <- rbind( Tot , body)
            rownames(z)[1] <- k[i]    
            rownames(z)[2] <- paste0("  ",rownames(ss1)[1])   
            
            fooz[[i]] <- z  
            
        } else {
          
          
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           # This step ensure if same SOC occurs more than once in a patient count only once
           sss <-unique(ss[,c("id","SOC","treatment")]) #  [1] drop PT
           Tot <- addmargins(table(sss$treatment))      #  [2]
          
           # this would allow duplicates counted within patients, 
           # not what is typically reported, code out [1], [2] and allow [3]
           # Tot <- ss1[nrow(ss1),]                      # [3]
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           
           sss <- unique(ss[,c("id","PT","treatment")]) #  [1] drop SOC
           ss1 <- addmargins(table(sss$PT, sss$treatment))
           
           body <- ss1[ c(1:(nrow(ss1)-1)),]   # select all rows except last
           z <- rbind( Tot , body)
           rownames(z)[1] <- k[i]    # add in SOC name
           # next line would add indent PTs, we do this later
           #rownames(z)[2:nrow(z)] <- paste0("  ",rownames(z)[2:nrow(z)])   
           fooz[[i]] <- z     # collect info
        
        }
    
       tmp <- do.call(rbind, fooz)      # bind the groups
                                    
       foo99 <- data.frame(tmp)  
  }
  

   
```

#  Build AE table

```{r, }   
  
   foo <- rbind(N0,N1,foo99)
   rownames(foo)[1] <- "Number of patients"
   rownames(foo)[2] <- "Number of patients with at least one adverse event"
   foo$Ap <- foo$Treatment.A/N0[1]*100
   foo$Bp <- foo$Treatment.B/N0[2]*100
   foo$Tp <- foo$Sum/N0[3]*100
   foo <- foo[,c("Treatment.A", "Ap","Treatment.B","Bp", "Sum","Tp")]
   foo$Ap <- formatz(foo$Ap);
   foo$Bp <- formatz(foo$Bp);
   foo$Tp <- formatz(foo$Tp);
   foo$Ap <- with(foo, paste0("(",  foo$Ap , ")"))
   foo$Bp <- with(foo, paste0("(",  foo$Bp , ")"))
   foo$Tp <- with(foo, paste0("(",  foo$Tp , ")"))
   names(foo) <- c("Treatment A", "(%)","Treatment B","(%)", "Total","(%)")

```

#  Print AE table

Except in the case of an SOC with one PT the sum of PTs may not necessarily equal the corresponding SOC. Investigate each SOC. One, within an SOC count how many patients have at least one of the particular SOC. Two, investigate all PTs within each SOC. Report how many patients have at least one. If the sum of PTs tally for an SOC, that means no patient has multiple PFs for that SOC.

Note also the 'Number of patients with at least one adverse event' is not expected to match the sum of the SOC, as a patient is likely to have more than one SOC.


```{r, }      
   
   pt <- grepl( "PT." ,  rownames(foo) ) # indent this term 
   
   d <- cbind("Primary System Organ Class / Preferred Term"= rownames(foo), foo)
              
   rownames(d) <- NULL
 
    # footnotes
    t1<- "- Adverse events are defined as new or worsening events that occur on or after the first study day of study treatment"
    t2<-"- A patient with multiple adverse events within a primary system organ class is counted only once for that system organ class"
    t3<-"- A patient with multiple adverse events within a preferred term is counted only once for that preferred term"
    t4<-"- System organ classes are presented in alphabetical order"
    t5<-"- Preferred terms are sorted within system organ class in alphabetical order"
    t6<-"- MedDRA version XY.Z has been used for reporting of adverse events"
    t7<-"- Study ABCMADEUPEFG; Database lock: 10Nov2020"

   
    tab_1 <-  kable(d ,
    format = "html", 
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
    caption = "Adverse events, by primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events, by primary system organ class and preferred term (Safety Analysis Set)" = 7) , bold = TRUE, font_size=15
    ) %>% footnote(
    general = c(t1,t2,t3,t4,t5,t6,t7),
    general_title = " ",
    footnote_as_chunk = F ,
    escape = F
  ) 
    
   add_indent(tab_1, which(pt), level_of_indent = 1)
   
   
```

#  Build AE table the wrong way

This will match the SOC simple tabulation further below

```{r, eval=TRUE}    
  
  
  call <- all[!all$count.AE %in% 0,]  # dispense with all obs with no AEs
  
  s <- call[, c("id","SOC","PT","treatment")]  # drop count vars
  
  #s <- unique(s)  # if within a patient has multiple SOC and linked PTs, 
                  # keep only 1 occurrence as this extra info is not used
  
  fooz<-list()                        # this will be used to collect data
  
  # analyse data in SOC chunks 

  k <- sort(unique(s$SOC))       # get a vector of unique SOC
  L <- length(k)                 # how many unique SOC 
  
  for (i in 1:L) {               # one by one analyze SOC groups
   
    ss <- s[s$SOC %in% k[i],]    # isolate a SOC chunk
    ss1 <- addmargins(table(ss$PT, ss$treatment))
       # special case if only 1 SOC row
    
        if (nrow(ss) %in% 1) {
          
           
            Tot <- ss1[rownames(ss1) %in% "Sum",]
            body <- ss1[!rownames(ss1) %in% "Sum",]
            z <- rbind( Tot , body)
            rownames(z)[1] <- k[i]    
            rownames(z)[2] <- paste0("  ",rownames(ss1)[1])   
            
            fooz[[i]] <- z  
            
        } else {
          
          
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           # This step ensure if same SOC occurs more than once in a patient count only once
           #sss <-unique(ss[,c("id","SOC","treatment")]) #  [1] drop PT
          # Tot <- addmargins(table(sss$treatment))      #  [2]
          
           # this would allow duplicates counted within patients, 
           # not what is typically reported, code out [1], [2] and allow [3]
           Tot <- ss1[nrow(ss1),]                      # [3]
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           
           sss <-  (ss[,c("PT","treatment")]) #  [1] drop SOC, note dropped unique
           ss1 <- addmargins(table(sss$PT, sss$treatment))
           
           body <- ss1[ c(1:(nrow(ss1)-1)),]   # select all rows except last
           z <- rbind( Tot , body)
           rownames(z)[1] <- k[i]    # add in SOC name
           # next line would add indent PTs, we do this later
           #rownames(z)[2:nrow(z)] <- paste0("  ",rownames(z)[2:nrow(z)])   
           fooz[[i]] <- z     # collect info
        
        }
    
       tmp <- do.call(rbind, fooz)      # bind the groups
                                    
       foo99 <- data.frame(tmp)  
  }
  
   #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
   foo <- rbind(N0,N1,foo99)
   rownames(foo)[1] <- "Number of patients"
   rownames(foo)[2] <- "Number of patients with at least one adverse event"
   foo$Ap <- foo$Treatment.A/N0[1]*100
   foo$Bp <- foo$Treatment.B/N0[2]*100
   foo$Tp <- foo$Sum/N0[3]*100
   foo <- foo[,c("Treatment.A", "Ap","Treatment.B","Bp", "Sum","Tp")]
   foo$Ap <- formatz(foo$Ap);
   foo$Bp <- formatz(foo$Bp);
   foo$Tp <- formatz(foo$Tp);
   foo$Ap <- with(foo, paste0("(",  foo$Ap , ")"))
   foo$Bp <- with(foo, paste0("(",  foo$Bp , ")"))
   foo$Tp <- with(foo, paste0("(",  foo$Tp , ")"))
   names(foo) <- c("Treatment A", "(%)","Treatment B","(%)", "Total","(%)")
 
   #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
   
   pt <- grepl( "PT." ,  rownames(foo) ) # indent this term 
   
   d <- cbind("Primary System Organ Class / Preferred Term"= rownames(foo), foo)
              
   rownames(d) <- NULL
 
    # footnotes
    t1<- "- Adverse events are defined as new or worsening events that occur on or after the first study day of study treatment"
    t2<-"- A patient with multiple adverse events within a primary system organ class is counted only once for that system organ class"
    t3<-"- A patient with multiple adverse events within a preferred term is counted only once for that preferred term"
    t4<-"- System organ classes are presented in alphabetical order"
    t5<-"- Preferred terms are sorted within system organ class in alphabetical order"
    t6<-"- MedDRA version XY.Z has been used for reporting of adverse events"
    t7<-"- Study ABCMADEUPEFG; Database lock: 10Nov2020"

   
    tab_1 <-  kable(d ,
    format = "html", 
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
    caption = "Adverse events, by primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events, by primary system organ class and preferred term (Safety Analysis Set)" = 7) , bold = TRUE, font_size=15
    ) %>% footnote(
    general = c(t1,t2,t3,t4,t5,t6,t7),
    general_title = " ",
    footnote_as_chunk = F ,
    escape = F
  ) 
    
   add_indent(tab_1, which(pt), level_of_indent = 1)
     

```

#  Word cloud SOC

```{r, }      
   
  # Create a vector containing only the text
  text <- all$SOC
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')

   
```

#  Word cloud preferred terms

```{r, }      
   
  # Create a vector containing only the text
  text <- all$PT
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')


```

#  The data

```{r, }   
   
 
  kable(all ,
    format = "html", 
    col.names = c("Patient ID", "Total AE", "AE count" , "System Organ Class", "Preferred Term", "Treatment") ,
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
  caption = "Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)" = 6) , bold = TRUE, font_size=15
    )  
  
```

#  Try Hmisc

This will straight out count SOCs, ignoring patient IDs and therefore multiple occurrences within patients , so this require further manipulation

```{r, }   


  
  f <- summaryM(SOC ~ treatment,   data=all,overall=TRUE,  test=FALSE )
  html(f)
  #plot(f)
  
```
 
Basic table that matches Hmisc

```{r, }     
  
  addmargins(table(all$SOC, all$treatment))
    
  # 
  # #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  #  ###patients with aleast one...look at title only
  # 
  #  atleast <- all[all$tot.AE %in% 1,]
  #  addmargins(table(atleast$treatment))
  #  f <- summaryM(SOC ~ treatment, test=TRUE, data=atleast, overall=TRUE)
  #  print(f)
  #  #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  #  
  #  
  #  uni <- unique(all[,c("SOC","PT","treatment")])
  #  
  #  # SOC headers
  #  f <- summaryM(SOC ~ treatment, test=TRUE, data=atleast, overall=TRUE)
  #  print(f)
  #  
  #  










```
# Computing Environment

`r markupSpecs$html$session()`