---
title: "AE reporting"
author: "Eamonn O'Brien<br><small><br>Biostatistics<br></small>"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    use_bookdown: true
    toc_depth: 3
    fig_caption: true
    code_folding: hide
link-citations: yes
description: "xxxxxx"
---
<style>
.table {
	width: 50%;
}
p.caption {
  font-size: 0.8em;
  color: DarkBlue;
}
body { 
	max-width: 1900px; 
	margin: 0 auto !important; 
	margin-right: 0 !important;
	float: none !important; 
}
</style>

```{css, echo=FALSE}

#https://designers.hubspot.com/docs/snippets/design/centering-your-website-using-max-width-and-auto-margins
#https://github.com/juba/rmdformats/issues/49
#https://stackoverflow.com/questions/60176526/kableextra-change-font-size-of-table-footnote
#https://towardsdatascience.com/create-a-word-cloud-with-r-bde3e7422e8a
#https://www.phusewiki.org/wiki/index.php?title=How_to_Count_Adverse_Events
#https://community.rstudio.com/t/dplyr-arrange-for-natural-sorting-of-several-columns/50028/2
 
```


```{r setup,echo=FALSE,include=FALSE}

  set.seed(234)
  library(rms)          # automatically engages rms(Hmisc)
  library(tidyverse)
  library(kableExtra) 
  library(knitr) 
  # wordcloud
  library(wordcloud)
  library(wordcloud2)
  library(RColorBrewer)
  library(tm)
  
  knitrSet(lang='markdown', w=7, h=7, dev='svg', fig.path='folder/')
  options(prType='html')
  options(width=200)
  
  # format function 
     formatz <- function(x){
        if (!is.na(x)  ) {
          sprintf(x, fmt = '%#.1f')  
         }
   }
 
```

# Introduction

Simple AE table example using kable in Rmarkdown

```{r, }

  m    <- 10      # m SOC categories we create, each will have m preferred terms
  n    <- 1000   # total sample size
  rate <- 0.33    # AE poisson rate
  J    <- 20      # maximum number of preferred terms that can occur

  
```

#  Make a fabricated medical dictionary

```{r, }

  require(stringi)
  require(word.alignment)

  myFun <- function(free="SOC.",n=1) { 
    
    paste0(free,paste(stri_rand_strings(sample(4:8,1),   # no of words
                                        sample(4:12,1)), # no of letters in words
                      #pattern = "[A-Za-z]",
                      collapse= " "))
  }
  
  N   <-  m

  x   <-  replicate(N, myFun()) # create N SOC groups
  p   <-  round(runif(N,2,J))   # size of SOC groups
  SOC <-  sort(rep(x, times=p)) # fill out the SOCs
  PT  <-  replicate(length(SOC), myFun(free="PT."))  # create unique Preferred terms 
  SOC <-  str_to_sentence(SOC)   # changes case so only first letter is uppercase
  PT <-   str_to_sentence(PT)    # changes case so only first letter is uppercase
  PT <- gsub("Pt","PT",PT)       # change case back
  SOC <- gsub("Soc","SOC",SOC)   # change case back
  db  <-  data.frame(cbind(SOC,PT))  # bind together, we now have the dictionary!   
  
##########################################################################################################
# make a second dictionary, maybe this is easier to explain the logic
  N   <- m # number of SOC
  S   <- rep(1:N) 
  p   <- round(runif(N,2,J))    
  d1 <-  sort(rep(S, times=p))  
  d1 <-  data.frame(cbind(id=d1))
  
  # count with group
  library(dplyr)
  d1 <- d1 %>%
    group_by(id) %>%
    mutate(count = seq(n()))
  
  SOC <- paste0("SOC",d1$id )
  # <- paste0("SOC",d1$id,"PT",d1$count)
  PT <-  paste0("PT",d1$count,".SOC",d1$id)
  db <-  data.frame(cbind(SOC, PT ))
##########################################################################################################

```

#  Create some AE data

```{r, }

 id <- 1:n             # patients
 ae <- rpois(n, rate)  # AEs per person
 
 zeroAE <- id[ae %in% 0]  # these ids have no AEs
 elements_2_remove <- zeroAE
 
 people.with.AE =   id[!(id %in% elements_2_remove)]
 people.with.no.AE = id[(id %in% elements_2_remove)]
 
 AE <- ae[!ae %in% 0]
 
 ex <- rep(people.with.AE, AE) 
 
 d1 <- data.frame(cbind(id=ex))
 d1 <- add_count(d1,id)
 
 d2 <- d1 %>%
   group_by(id, n) %>%
   mutate(count = seq(n()))

 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~#people with no AEs
 people.with.no.AE = id[(id %in% elements_2_remove)]
 
 AE <- ae[ae %in% 0]
 
 d1 <- data.frame(cbind(id=people.with.no.AE, n=0,count=0))
 
 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 all <- rbind(d1,d2)
 
 all <- plyr::arrange(all, id, count)
 
 names(all) <- c("id","tot.AE","count.AE")

 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  AES <- db[sample(nrow(db), nrow(d2), replace=TRUE), ]
 
  d2 <- cbind(d2, AES)
 
 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  d1$SOC="";d1$PT=""
  
  all <- rbind(d1,d2)
  
  all <- plyr::arrange(all, id, count)
  
  names(all) <- c("id","tot.AE","count.AE","SOC","PT")
 
  #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  # add trt
  trt <- sample(c("Treatment A","Treatment B"), n, replace=TRUE)
  
  all$treatment <- rep(trt, times=as.vector(table(all$id)))
  all$treatment <- factor(all$treatment)

```

#  Calculate N

```{r, }  
  
  N <- unique(all[, c("id", "treatment")])
  N0 <- addmargins(table(N$treatment))
 
```

# Calculate the number of patients with at least one AE

```{r, }    
  
  N <- unique(all[, c("id", "tot.AE","treatment")])
  atleast <- N[!N$tot.AE %in% 0,]
  N1 <- addmargins(table(atleast$treatment))

```

# Manage data for by System order class : Preferred terms, the right way

```{r, eval=TRUE}  
 
  
  call <- all[!all$count.AE %in% 0,]  # dispense with all obs with no AEs
  
  s <- call[, c("id","SOC","PT","treatment")]  # drop count vars
  
 # s <- unique(s)  # if within a patient has multiple SOC and linked PTs, 
                  # keep only 1 occurrence as this extra info is not used
                  # this will not make a difference as I deal with this in the following code
  #s <- s[ gtools::mixedsort(s$SOC), ]  # sort alphanumeric   all3 <- all3[gtools::mixedorder(all3$PT), ]
 
  # s<-s[!(s$SOC=="" ),]  # no need for this as 
  #s <- s[gtools::mixedorder(s$SOC), ]  # sort before manipulation
  
  fooz<-list()                        # this will be used to collect data
  
  # analyse data in SOC chunks 

  k <- sort(unique(s$SOC))       # get a vector of unique SOC
  k <- gtools::mixedsort(unique(s$SOC)) 
  L <- length(k)                 # how many unique SOC 
  
  for (i in 1:L) {               # one by one analyze SOC groups
   
    ss <- s[s$SOC %in% k[i],]    # isolate a SOC chunk
 
       # special case if only 1 SOC row
        if (nrow(ss) %in% 1) {
          
            ss1 <- addmargins(table(ss$PT, ss$treatment))
            Tot <- ss1[rownames(ss1) %in% "Sum",]
            body <- ss1[!rownames(ss1) %in% "Sum",]
            z <- rbind( Tot , body)
            rownames(z)[1] <- k[i]    
            rownames(z)[2] <- paste0("  ",rownames(ss1)[1])   
            
            fooz[[i]] <- z  
            
        } else {
          
          
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           # This step ensure if same SOC occurs more than once in a patient count only once
           sss <-unique(ss[,c("id","SOC","treatment")]) #  [1] drop PT
           Tot <- addmargins(table(sss$treatment))      #  [2]
          
           # this would allow duplicates counted within patients, 
           # not what is typically reported, code out [1], [2] and allow [3]
           # Tot <- ss1[nrow(ss1),]                      # [3]
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           
           sss <- unique(ss[,c("id","PT","treatment")]) #  [1] drop SOC
           ss1 <- addmargins(table(sss$PT, sss$treatment))
           
           body <- ss1[ c(1:(nrow(ss1)-1)),]   # select all rows except last
           body <- body[ gtools::mixedsort(row.names(body)), ]  # sort alphanumeric
           z <- rbind( Tot , body)
          
           
           rownames(z)[1] <- k[i]    # add in SOC name
           # next line would add indent PTs, we do this later
           #rownames(z)[2:nrow(z)] <- paste0("  ",rownames(z)[2:nrow(z)])   
           
           fooz[[i]] <- z     # collect info
        
        }
    
       tmp <- do.call(rbind, fooz)      # bind the groups
                                    
       foo99 <- data.frame(tmp)  
  }
  

   
```

#  Build AE table

```{r, }   
  
   foo <- rbind(N0,N1,foo99)
   rownames(foo)[1] <- "Number of patients"
   rownames(foo)[2] <- "Number of patients with at least one adverse event"
   foo$Ap <- foo$Treatment.A/N0[1]*100
   foo$Bp <- foo$Treatment.B/N0[2]*100
   foo$Tp <- foo$Sum/N0[3]*100
   foo <- foo[,c("Treatment.A", "Ap","Treatment.B","Bp", "Sum","Tp")]
   foo$Ap <- formatz(foo$Ap);
   foo$Bp <- formatz(foo$Bp);
   foo$Tp <- formatz(foo$Tp);
   foo$Ap <- with(foo, paste0("(",  foo$Ap , ")"))
   foo$Bp <- with(foo, paste0("(",  foo$Bp , ")"))
   foo$Tp <- with(foo, paste0("(",  foo$Tp , ")"))
   names(foo) <- c("Treatment A", "(%)","Treatment B","(%)", "Total","(%)")

```

#  Print AE table

Each patient is counted only once within each category. So when one patient has two headache AEs, then this patient is counted once under headache. Patient are also only counted once within each body system, so when a patient is having different AEs within the same body system. This means, that the numbers of the preferred terms does not sum up to the body system numbers. 

Except in the case of an SOC with one PT the sum of PTs may not necessarily equal the corresponding SOC. Investigate each SOC. One, within an SOC count how many patients have at least one of the particular SOC. Two, investigate all PTs within each SOC. Report how many patients have at least one. If the sum of PTs tally for an SOC, that means no patient has multiple PFs for that SOC.

Note also the 'Number of patients with at least one adverse event' is not expected to match the sum of the SOC, as a patient is likely to have more than one SOC.

Ways of saying the same thing: 'Patients with more than one occurrence of a preferred term are counted only once.'
'A patient with multiple occurrences of an AE is counted only once in the AE category.'

'Patients with any adverse event' is synonymous with 'Patients with at least one AE'

```{r, }      
   
   pt <- grepl( "PT" ,  rownames(foo) ) # indent this term 
   
   d <- cbind("Primary System Organ Class / Preferred Term"= rownames(foo), foo)
              
   rownames(d) <- NULL
 
    # footnotes
    t1<- "- Adverse events are defined as new or worsening events that occur on or after the first study day of study treatment"
    t2<-"- A patient with multiple adverse events within a primary system organ class is counted only once for that system organ class"
    t3<-"- A patient with multiple adverse events within a preferred term is counted only once for that preferred term"
    t3a<-"- A patient with multiple occurrences of an AE is counted only once in the AE category"
    t4<-"- System organ classes are presented in alphabetical order"
    t5<-"- Preferred terms are sorted within system organ class in alphabetical order"
    t6<-"- MedDRA version XY.Z has been used for reporting of adverse events"
    t7<-"- Study ABCMADEUPEFG; Database lock: 10Nov2020"

   
    tab_1 <-  kable(d ,
    format = "html", 
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
    caption = "Adverse events, by primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events, by primary system organ class and preferred term (Safety Analysis Set)" = 7) , bold = TRUE, font_size=15
    ) %>% footnote(
    general = c(t1, t3a,t4,t5,t6,t7),
    general_title = " ",
    footnote_as_chunk = F ,
    escape = F
  ) 
    
   add_indent(tab_1, which(pt), level_of_indent = 1)
   
   
```

#  Build AE table the wrong way

This will match the SOC simple tabulation further below. It may coincidentally match the previous correct table, especially if the sample is small

```{r, eval=TRUE}    
  
  
  call <- all[!all$count.AE %in% 0,]  # dispense with all obs with no AEs
  
  s <- call[, c("id","SOC","PT","treatment")]  # drop count vars
  
  #s <- unique(s)  # if within a patient has multiple SOC and linked PTs, 
                  # keep only 1 occurrence as this extra info is not used
  
  fooz<-list()                        # this will be used to collect data
  
  # analyse data in SOC chunks 

 # k <- sort(unique(s$SOC))       # get a vector of unique SOC
  k <- gtools::mixedsort(unique(s$SOC)) 
  L <- length(k)                 # how many unique SOC 
  
  for (i in 1:L) {               # one by one analyze SOC groups
   
    ss <- s[s$SOC %in% k[i],]    # isolate a SOC chunk
    ss1 <- addmargins(table(ss$PT, ss$treatment))
       # special case if only 1 SOC row
    
        if (nrow(ss) %in% 1) {
          
           
            Tot <- ss1[rownames(ss1) %in% "Sum",]
            body <- ss1[!rownames(ss1) %in% "Sum",]
            z <- rbind( Tot , body)
            rownames(z)[1] <- k[i]    
            rownames(z)[2] <- paste0("  ",rownames(ss1)[1])   
            
            fooz[[i]] <- z  
            
        } else {
          
          
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           # This step ensure if same SOC occurs more than once in a patient count only once
           #sss <-unique(ss[,c("id","SOC","treatment")]) #  [1] drop PT
          # Tot <- addmargins(table(sss$treatment))      #  [2]
          
           # this would allow duplicates counted within patients, 
           # not what is typically reported, code out [1], [2] and allow [3]
           Tot <- ss1[nrow(ss1),]                      # [3]
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           
           sss <-  (ss[,c("PT","treatment")]) #  [1] drop SOC, note dropped unique
           ss1 <- addmargins(table(sss$PT, sss$treatment))
           
           body <- ss1[ c(1:(nrow(ss1)-1)),]   # select all rows except last
           body <- body[ gtools::mixedsort(row.names(body)), ]  # sort alphanumeric

           z <- rbind( Tot , body)
           rownames(z)[1] <- k[i]    # add in SOC name
           # next line would add indent PTs, we do this later
           #rownames(z)[2:nrow(z)] <- paste0("  ",rownames(z)[2:nrow(z)])   
           fooz[[i]] <- z     # collect info
        
        }
    
       tmp <- do.call(rbind, fooz)      # bind the groups
                                    
       foo99 <- data.frame(tmp)  
  }
  
   #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
   foo <- rbind(N0,N1,foo99)
   rownames(foo)[1] <- "Number of patients"
   rownames(foo)[2] <- "Number of patients with at least one adverse event"
   foo$Ap <- foo$Treatment.A/N0[1]*100
   foo$Bp <- foo$Treatment.B/N0[2]*100
   foo$Tp <- foo$Sum/N0[3]*100
   foo <- foo[,c("Treatment.A", "Ap","Treatment.B","Bp", "Sum","Tp")]
   foo$Ap <- formatz(foo$Ap);
   foo$Bp <- formatz(foo$Bp);
   foo$Tp <- formatz(foo$Tp);
   foo$Ap <- with(foo, paste0("(",  foo$Ap , ")"))
   foo$Bp <- with(foo, paste0("(",  foo$Bp , ")"))
   foo$Tp <- with(foo, paste0("(",  foo$Tp , ")"))
   names(foo) <- c("Treatment A", "(%)","Treatment B","(%)", "Total","(%)")
 
   #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
   
   pt <- grepl( "PT." ,  rownames(foo) ) # indent this term 
   
   d <- cbind("Primary System Organ Class / Preferred Term"= rownames(foo), foo)
              
   rownames(d) <- NULL
 
    # footnotes
    # t1<- "- Adverse events are defined as new or worsening events that occur on or after the first study day of study treatment"
    # t2<-"- A patient with multiple adverse events within a primary system organ class is counted only once for that system organ class"
    # t3<-"- A patient with multiple adverse events within a preferred term is counted only once for that preferred term"
    # t4<-"- System organ classes are presented in alphabetical order"
    # t5<-"- Preferred terms are sorted within system organ class in alphabetical order"
    # t6<-"- MedDRA version XY.Z has been used for reporting of adverse events"
    # t7<-"- Study ABCMADEUPEFG; Database lock: 10Nov2020"

   
    tab_1 <-  kable(d ,
    format = "html", 
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
    caption = "Adverse events, by primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events, by primary system organ class and preferred term (Safety Analysis Set)" = 7) , bold = TRUE, font_size=15
    ) %>% footnote(
    general = c(t1,t2,t3a,t5,t6,t7),
    general_title = " ",
    footnote_as_chunk = F ,
    escape = F
  ) 
    
   add_indent(tab_1, which(pt), level_of_indent = 1)
     

```

#  Word cloud SOC, (to do ensure this is based on counts in table)

```{r, }      
   
  # Create a vector containing only the text
  text <- all$SOC
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')

   
```

#  Word cloud preferred terms (to do ensure this is based on counts in table)

```{r, }      
   
  # Create a vector containing only the text
  text <- all$PT
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')


```

#  The data

```{r, }   
   
 
  kable(all ,
    format = "html", 
    col.names = c("Patient ID", "Total AE", "AE count" , "System Organ Class", "Preferred Term", "Treatment") ,
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
  caption = "Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)" = 6) , bold = TRUE, font_size=15
    )  


```

#  Are there duplicates of SOC within patient?

```{r, }   
   
  # duplicate PT within a person
  # all2 <- all[!all$tot.AE %in% 0,] # remove patients with no AE
  # all3 <- arrange(all2, SOC, treatment , id)
  # chk <- all3[, c("id","PT")]
  # # ids of duplication
  # ids <- all3[which(duplicated(chk)),]$id
  # all3 <- all3[gtools::mixedorder(all3$PT), ]
  # all3[all3$id %in% ids,]
  
  require(gtools)
   
  all2 <- all[!all$tot.AE %in% 0,] #  remove patients with no AE

  all2 <- arrange(all2, SOC, PT,id, treatment)
  
  # https://community.rstudio.com/t/filter-out-all-rows-with-duplicate-values/41043/4
  # flag duplicates
  all2 = all2 %>% 
  group_by(id, SOC) %>% 
  mutate(duplicate.soc.flag = n() > 1)
  
  all2$duplicate.soc.flag<-all2$duplicate.soc.flag*1
  
  all2 = all2 %>% 
  group_by(id, PT) %>% 
  mutate(duplicate.pt.flag = n() > 1)
  
  all2$duplicate.pt.flag<-all2$duplicate.pt.flag*1
  
  all2 <- all2[mixedorder(all2$SOC),]
  
  all2[all2$duplicate.soc.flag %in%  1,]
  
  

```

#  Are there duplicates of PT within patient?

```{r, }     
  
  # duplicated SOC within a person
  # 
  # chk <- all3[, c("id","SOC")]
  # # ids of duplication
  # ids <- all3[which(duplicated(chk)),]$id 
  # all3 <- all3[gtools::mixedorder(all3$SOC), ]
  # 
  # all3[all3$id %in% ids,]
  all2[all2$duplicate.pt.flag %in%  1,]
  

```

# Use datatable package for listing with better interactivity

```{r,}

 DT::datatable(all2, rownames=FALSE,
               plugins = 'natural',
               options = list(
 # dom = 't',
  columnDefs = list(list(type = 'natural', targets = c(4,5)))
))
  
#DT::datatable(all2, rownames=FALSE)

# datatable(d, plugins = 'natural', options = list(
#   dom = 't',
#   columnDefs = list(list(type = 'natural', targets = 6))
# ))


```

#  The data again removing no AE patients and sorted

```{r, }     

 require(gtools)
   
  all2 <- all[!all$tot.AE %in% 0,] #  remove patients with no AE

  all2 <- arrange(all2, SOC, PT,id, treatment)
  
  # https://community.rstudio.com/t/filter-out-all-rows-with-duplicate-values/41043/4
  # flag duplicates
  all2 = all2 %>% 
  group_by(id, SOC) %>% 
  mutate(duplicate.soc.flag = n() > 1)
  
  all2$duplicate.soc.flag<-all2$duplicate.soc.flag*1
  
  all2 = all2 %>% 
  group_by(id, PT) %>% 
  mutate(duplicate.pt.flag = n() > 1)
  
  all2$duplicate.pt.flag<-all2$duplicate.pt.flag*1
  
  # all2 <- all2[c(mixedorder(all2$SOC), mixedorder(all2$PT)),]

    all2 <- all2[mixedorder(all2$SOC),]
   
  kable(all2 ,
    format = "html", 
    col.names = c("Patient ID", "Total AE", "AE count" , "System Organ Class", "Preferred Term", "Treatment", "SOC dup?","PT dup?") ,
    table.attr = "style='width:80%;'",
    align = "lcccccccc" , 
  caption = "Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)" = 8) , bold = TRUE, font_size=15
    ) 
  
  
  
```

#  Try Hmisc

This will straight out count SOCs, ignoring patient IDs and therefore multiple occurrences within patients , so this require further manipulation

```{r, }   


  
  f <- summaryM(SOC ~ treatment,   data=all,overall=TRUE,  test=FALSE )
  html(f)
  #plot(f)
  
```
 
# Basic table that matches Hmisc

```{r, }     
  
  f <- addmargins(table(all$SOC, all$treatment))
    
 #  f <- f[ gtools::mixedsort(row.names(f)), ]  
f
  # 
  # #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  #  ###patients with aleast one...look at title only
  # 
  #  atleast <- all[all$tot.AE %in% 1,]
  #  addmargins(table(atleast$treatment))
  #  f <- summaryM(SOC ~ treatment, test=TRUE, data=atleast, overall=TRUE)
  #  print(f)
  #  #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  #  
  #  
  #  uni <- unique(all[,c("SOC","PT","treatment")])
  #  
  #  # SOC headers
  #  f <- summaryM(SOC ~ treatment, test=TRUE, data=atleast, overall=TRUE)
  #  print(f)
  #  
  #  

# Example showing 
# Data				Summary		
# ID						
# 1	SOC1	PT1			SOC1	2
# 1	SOC1	PT1			PT1	1
# 1	SOC1	PT2			PT2	1
# 2	SOC1	PT4			PT4	1
# 2	SCO2	PT5			SCO2	1
# 3	SOC3	PT7			PT5	1
# 3	SOC3	PT8			SOC3	3
# 4	SOC3	PT7			PT7	2
# 5	SOC3	PT9			PT8	1
# 				      	PT9	1

```

# Manage data for by System order class : Preferred terms, the right way for Word Cloud

```{r, eval=TRUE}  
 
  
  call <- all[!all$count.AE %in% 0,]  # dispense with all obs with no AEs
  
  s <- call[, c("id","SOC","PT","treatment")]  # drop count vars

  fooz<-list()                        # this will be used to collect data
  
  # analyse data in SOC chunks 

  k <- sort(unique(s$SOC))       # get a vector of unique SOC
  k <- gtools::mixedsort(unique(s$SOC)) 
  L <- length(k)                 # how many unique SOC 
  
  for (i in 1:L) {               # one by one analyze SOC groups
   
    ss <- s[s$SOC %in% k[i],]    # isolate a SOC chunk
 
       # special case if only 1 SOC row
        if (nrow(ss) %in% 1) {
          
            ss1 <- addmargins(table(ss$PT, ss$treatment))
            Tot <- ss1[rownames(ss1) %in% "Sum",]
          
            z <- ( Tot )

            fooz[[i]] <- z  
            
        } else {
          
  #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           # This step ensure if same SOC occurs more than once in a patient count only once
           sss <-unique(ss[,c("id","SOC","treatment")]) #  [1] drop PT
           Tot <- addmargins(table(sss$treatment))      #  [2]
          
           z <-  ( Tot  )
          
           rownames(z)[1] <- k[i]    # add in SOC name

           
           fooz[[i]] <- z     # collect info
        
        }
    
       tmp <- do.call(rbind, fooz)      # bind the groups
                                    
       foo99s <- data.frame(tmp)  
      
  }
  
     rownames(foo99s) <- k      
   arrange(foo99s, Sum) 

```

# Manage data for by System order class : Preferred terms, the right way for word cloud

```{r, eval=TRUE}  
 
  
  call <- all[!all$count.AE %in% 0,]  # dispense with all obs with no AEs
  
  s <- call[, c("id","SOC","PT","treatment")]  # drop count vars
  
 
  
  fooz<-list()                        # this will be used to collect data
  
  # get a vector of unique SOC
  k <- gtools::mixedsort(unique(s$SOC)) 
  L <- length(k)                 # how many unique SOC 
  
  for (i in 1:L) {               # one by one analyze SOC groups
   
    ss <- s[s$SOC %in% k[i],]    # isolate a SOC chunk
 
       # special case if only 1 SOC row
        if (nrow(ss) %in% 1) {
          
            ss1 <- addmargins(table(ss$PT, ss$treatment))
        #    Tot <- ss1[rownames(ss1) %in% "Sum",]
            body <- ss1[!rownames(ss1) %in% "Sum",]
            z <- (body)
         #   rownames(z)[1] <- k[i]    
            rownames(z)[2] <- paste0("  ",rownames(ss1)[1])   
            
            fooz[[i]] <- z  
            
        } else {
          
          
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           # This step ensure if same SOC occurs more than once in a patient count only once
         #  sss <-unique(ss[,c("id","SOC","treatment")]) #  [1] drop PT
          # Tot <- addmargins(table(sss$treatment))      #  [2]
          
           # this would allow duplicates counted within patients, 
           # not what is typically reported, code out [1], [2] and allow [3]
           # Tot <- ss1[nrow(ss1),]                      # [3]
           #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
           
           sss <- unique(ss[,c("id","PT","treatment")]) #  [1] drop SOC
           ss1 <- addmargins(table(sss$PT, sss$treatment))
           
           body <- ss1[ c(1:(nrow(ss1)-1)),]   # select all rows except last
           body <- body[ gtools::mixedsort(row.names(body)), ]  # sort alphanumeric
           z <- rbind( body)
          
           
       #    rownames(z)[1] <- k[i]    # add in SOC name
           # next line would add indent PTs, we do this later
           #rownames(z)[2:nrow(z)] <- paste0("  ",rownames(z)[2:nrow(z)])   
           
           fooz[[i]] <- z     # collect info
        
        }
    
       tmp <- do.call(rbind, fooz)      # bind the groups
                                    
       foo99p <- data.frame(tmp)  
  }
  
     arrange(foo99p, Sum) 
   

  
```

#  Word cloud SOC, ( this is based on counts in correct table)

```{r, }      
   
  # Create a vector containing only the text
  text <- rep(rownames(foo99s),foo99s$Sum)
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')

   
```

#  Word cloud preferred terms (this is based on counts in correct table)

```{r, }      
   
  # Create a vector containing only the text
  text <- rep(rownames(foo99p),foo99p$Sum)
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')
 
  
     
```

# Computing Environment

`r markupSpecs$html$session()`