---
title: "AE reporting"
author: "Eamonn O'Brien<br><small><br>Department of Insanity<br></small>"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    use_bookdown: true
    toc_depth: 3
    fig_caption: true
    code_folding: hide
link-citations: yes
description: "xxxxxx"
---
<style>
.table {
	width: 50%;
}
p.caption {
  font-size: 0.8em;
  color: DarkBlue;
}
body { 
	max-width: 1900px; 
	margin: 0 auto !important; 
	margin-right: 0 !important;
	float: none !important; 
}
</style>

```{css, echo=FALSE}

#https://designers.hubspot.com/docs/snippets/design/centering-your-website-using-max-width-and-auto-margins
#https://github.com/juba/rmdformats/issues/49
#https://stackoverflow.com/questions/60176526/kableextra-change-font-size-of-table-footnote
#https://towardsdatascience.com/create-a-word-cloud-with-r-bde3e7422e8a
 
```


```{r setup,echo=FALSE,include=FALSE}

  set.seed(234)
  library(rms)          # automatically engages rms(Hmisc)
  library(tidyverse)
  library(kableExtra) 
  library(knitr) 
  # wordcloud
  library(wordcloud)
  library(wordcloud2)
  library(RColorBrewer)
  library(tm)
  
  knitrSet(lang='markdown', w=7, h=7, dev='svg', fig.path='folder/')
  options(prType='html')
  options(width=200)
  
  # format function 
     formatz <- function(x){
        if (!is.na(x)  ) {
          sprintf(x, fmt = '%#.1f')  
         }
   }
 
```

# Introduction

Simple AE table example using kable in Rmarkdown

```{r, }

  m    <- 10      # m SOC categories we create, each will have m preferred terms
  n    <- 150     # total sample size
  rate <- 0.33    # AE poisson rate
  J    <- 20      # maximum number of preferred terms that can occur

  
```

#  Make a fabricated medical dictionary

```{r, }

  require(stringi)

  myFun <- function(free="SOC.",n=1) { 
    
    paste0(free,paste(stri_rand_strings(sample(2:5,1), 
                                        sample(2:10, n, replace=TRUE)), 
                      collapse= " "))
  }
  
  N   <-  m
  x   <-  replicate(N, myFun())
  p   <-  round(runif(N,2,J))   
  SOC <-  sort(rep(x, times=p)) 
  PT  <-  replicate(length(SOC), myFun(free="PT."))
  db  <-  data.frame(cbind(SOC,PT))

  
```

#  Create some AE data

```{r, }

 id <- 1:n             # patients
 ae <- rpois(n, rate)  # AEs per person
 
 zeroAE <- id[ae %in% 0]  # these ids have no AEs
 elements_2_remove <- zeroAE
 
 people.with.AE =   id[!(id %in% elements_2_remove)]
 people.with.no.AE = id[(id %in% elements_2_remove)]
 
 AE <- ae[!ae %in% 0]
 
 ex <- rep(people.with.AE, AE) 
 
 d1 <- data.frame(cbind(id=ex))
 d1 <- add_count(d1,id)
 
 d2 <- d1 %>%
   group_by(id, n) %>%
   mutate(count = seq(n()))

 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~#people with no AEs
 people.with.no.AE = id[(id %in% elements_2_remove)]
 
 AE <- ae[ae %in% 0]
 
 d1 <- data.frame(cbind(id=people.with.no.AE, n=0,count=0))
 
 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 all <- rbind(d1,d2)
 
 all <- plyr::arrange(all, id, count)
 
 names(all) <- c("id","tot.AE","count.AE")

 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  AES <- db[sample(nrow(db), nrow(d2), replace=TRUE), ]
 
  d2 <- cbind(d2, AES)
 
 #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  d1$SOC="";d1$PT=""
  
  all <- rbind(d1,d2)
  
  all <- plyr::arrange(all, id, count)
  
  names(all) <- c("id","tot.AE","count.AE","SOC","PT")
 
  #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  # add trt
  trt <- sample(c("Treatment A","Treatment B"), n, replace=TRUE)
  
  all$treatment <- rep(trt, times=as.vector(table(all$id)))
  all$treatment <- factor(all$treatment)

```

#  Calculate N

```{r, }  
  
  N <- unique(all[, c("id", "treatment")])
  N0 <- addmargins(table(N$treatment))
 
```

# Calculate the number of patients with at least one AE

```{r, }    
  
  N <- unique(all[, c("id", "tot.AE","treatment")])
  atleast <- N[!N$tot.AE %in% 0,]
  N1 <- addmargins(table(atleast$treatment))

```

# Manage data for by System order class : Preferred terms

```{r, }  

  fooz<-list()  # this will be used to collect data
  # PTs within each SOC
  s <- all[, c("id","SOC","PT","treatment")]
  s <- unique(s)
  s<-s[!(s$SOC=="" ),]
  
  k <- sort(unique(s$SOC))
  L <- length(k)
  for (i in 1:L) {
   
    ss <- s[s$SOC %in% k[i],]
    
    ss1 <- addmargins(table(ss$PT, ss$treatment))
    
        if (nrow(ss) %in% 1) {
            Tot <- ss1[rownames(ss1) %in% "Sum",]
            body <- ss1[!rownames(ss1) %in% "Sum",]
            z <- rbind( Tot , body)
            rownames(z)[1] <- k[i]    
            rownames(z)[2] <- paste0("  ",rownames(ss1)[1])    
            fooz[[i]]<-z  
            
        } else {
    
           # put SOC at top and indent PTs
           Tot <- ss1[nrow(ss1),]
           # select all rows except last
           body <- ss1[ c(1:(nrow(ss1)-1)),]
           z <- rbind( Tot , body)
           rownames(z)[1] <- k[i]    
           rownames(z)[2:nrow(z)] <- paste0("  ",rownames(z)[2:nrow(z)])   
           fooz[[i]]<-z  
        
        }
    
       tmp <- do.call(rbind, fooz)      # bind the groups
                                    
       foo99 <- data.frame(tmp)  
  }

```

#  Build AE table

```{r, }   
  
   foo <- rbind(N0,N1,foo99)
   rownames(foo)[1] <- "Number of patients"
   rownames(foo)[2] <- "Number of patients with at least one adverse event"
   foo$Ap <- foo$Treatment.A/N0[1]*100
   foo$Bp <- foo$Treatment.B/N0[2]*100
   foo$Tp <- foo$Sum/N0[3]*100
   foo <- foo[,c("Treatment.A", "Ap","Treatment.B","Bp", "Sum","Tp")]
   foo$Ap <- formatz(foo$Ap);
   foo$Bp <- formatz(foo$Bp);
   foo$Tp <- formatz(foo$Tp);
   foo$Ap <- with(foo, paste0("(",  foo$Ap , ")"))
   foo$Bp <- with(foo, paste0("(",  foo$Bp , ")"))
   foo$Tp <- with(foo, paste0("(",  foo$Tp , ")"))
   names(foo) <- c("Treatment A", "(%)","Treatment B","(%)", "Total","(%)")

```

#  Print AE table

```{r, }      
   
   pt <- grepl( "PT." ,  rownames(foo) )
   
   d <- cbind("Primary System Organ Class / Preferred Term"= rownames(foo), foo)
              
   rownames(d) <- NULL
 
    # footnotes
    t1<- "- Adverse events are defined as new or worsening events that occur on or after the first study day of study treatment"
    t2<-"- A patient with multiple adverse events within a primary system organ class is counted only once"
    t3<-"- System organ classes are presented in alphabetical order"
    t4<-"- Preferred terms are sorted within system organ class in alphabetical order"
    t5<-"- MedDRA version XY.Z has been used for reporting of adverse events"
    t6<-"- Study ABCMADEUPEFG; Database lock: 10Nov2020"

   
    tab_1 <-  kable(d ,
    format = "html", 
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
    caption = "Adverse events, by primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events, by primary system organ class and preferred term (Safety Analysis Set)" = 7) , bold = TRUE, font_size=15
    ) %>% footnote(
    general = c(t1,t2,t3,t4,t5,t6),
    general_title = " ",
    footnote_as_chunk = F ,
    escape = F
  ) 
    
   add_indent(tab_1, which(pt), level_of_indent = 1)

```

#  Word cloud SOC

```{r, }      
   
  # Create a vector containing only the text
  text <- all$SOC
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')

   
```

#  Word cloud preferred terms

```{r, }      
   
  # Create a vector containing only the text
  text <- all$PT
  text <- text [! text %in% ""]

  # Create a corpus  
  docs <- Corpus(VectorSource(text))
  dtm <- TermDocumentMatrix(docs) 
  matrix <- as.matrix(dtm) 
  words <- sort(rowSums(matrix),decreasing=TRUE) 
  df <- data.frame(word = names(words),freq=words)

  wordcloud(words = df$word, freq = df$freq, min.freq = 1,
            max.words=200, random.order=FALSE, rot.per=0.35,
            colors=brewer.pal(8, "Dark2"))

#  wordcloud2(data=df, size=1.6, color='random-dark')


```

#  The data

```{r, }   
   
 
  kable(all ,
    format = "html", 
    col.names = c("Patient ID", "Total AE", "AE count" , "System Organ Class", "Preferred Term", "Treatment") ,
    table.attr = "style='width:80%;'",
    align = "lcccccc" , 
  caption = "Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)",
    escape = F
  ) %>%
  kable_styling(
    full_width = T ,
    bootstrap_options = c("hover", "condensed" , "bordered"),
    font_size = 14,
    position = "center", 
  ) %>% row_spec(0, font_size=14
  ) %>% add_header_above(c("Adverse events listing, primary system organ class and preferred term (Safety Analysis Set)" = 6) , bold = TRUE, font_size=15
    )  
  
```

#  Try Hmisc

This will straight out count SOCs, ignoring patient IDs, and multiple occurrences within patients So will need to do further manipulation

```{r, }   


  
  f <- summaryM(SOC ~ treatment,   data=all,overall=TRUE,  test=FALSE )
  html(f)
  #plot(f)
  
```
 
Basic table that matches Hmisc

```{r, }     
  
  addmargins(table(all$SOC, all$treatment))
    
  # 
  # #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  #  ###patients with aleast one...look at title only
  # 
  #  atleast <- all[all$tot.AE %in% 1,]
  #  addmargins(table(atleast$treatment))
  #  f <- summaryM(SOC ~ treatment, test=TRUE, data=atleast, overall=TRUE)
  #  print(f)
  #  #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
  #  
  #  
  #  uni <- unique(all[,c("SOC","PT","treatment")])
  #  
  #  # SOC headers
  #  f <- summaryM(SOC ~ treatment, test=TRUE, data=atleast, overall=TRUE)
  #  print(f)
  #  
  #  










```
# Computing Environment

`r markupSpecs$html$session()`